---
title: 'Final Project - Step 2 (15 Points)'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author:
    affiliations:
      - name: "Spring 2025"
affiliation-title: "Quarter"
format: 
 pdf: 
    latex_engine: xelatex
    geometry: "letterpaper, margin=0.3in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```



:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- Valerie De La Fuente    (valeriedelafuente)
- Matthew Arteaga    (matthewarteaga)
- STUDENT 3    (NetID 3)
- STUDENT 4    (NetID 4)
- STUDENT 5    (NetID 5)
:::


::: {.callout-caution}
## Due Date
The deadline for this step is **Friday, May 9, 2025**.
:::


:::{.callout-tip}
# Instructions
In this step, you will develop clear research questions and hypotheses based on your selected dataset, and conduct a thorough Exploratory Data Analysis (EDA). This foundational work is crucial for guiding your analysis in the following steps.

:::

# Step 2: Research Questions, Hypotheses, and Exploratory Data Analysis (EDA)


## Research Questions

**Question 1**

Do certain dietary habits coincide with an increased rate of depression among students?

**Question 2**
Is there a correlation between the amount of sleep a student gets and the proportion of them that are depressed?

**Question 3**

Does the presence (and magnitude) of certain stressors have an impact on the rate at which students are depressed?

## Hypotheses

**Hypothesis 1**

Students with moderate to healthy dietary habits will have lower rates of depression compared to students with unhealthy dietary habits.

**Hypothesis 2**

Students who average more sleep per night will have lower rates of depression compared to students who average less.

**Hypothesis 3**

Students with the highest collective reported stressors (`Academic Pressure + Work Pressure + Financial Stress`) will have higher rates of depression compared to students with lower collective reported stressors.

## Exploratory Data Analysis (EDA)

## Data Cleaning

```{r}
# Load necessary packages
library(readr)
library(tidyverse)

# Load in the data
# depression_data <- read_csv("data/student_depression_dataset.csv")
depression_data <- read_csv("student_depression_dataset.csv")

# View the dataset
head(depression_data)
```


## Descriptive Statistics


## Data Visualization

