---
title: 'Mini Project 2'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author: "Instructor: Ali Abuzaid"
format: 
 pdf:
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```

:::: callout
::: {style="text-align: center"}
[**STUDENT NAME**]{style="color: blue;"}
:::

-   Phuc Lu (pdlu)
-   Valerie De La Fuente (valeriedelafuente)
::::

::: callout-tip
# Instructions

-   This mini project is designed to give you practical experience with real-world data using R and Shiny. You’ll create an interactive web application that allows users to explore and visualize a dataset.

-   Work in groups of **2 students** from the same discussion section.

-   Individual submissions will not be accepted.

-   Please use the provided `MP 2.qmd` file to type your Documentation and Presentation and submit it as a PDF file. You can utilize `RStudio`for this purpose. For guidance, refer to the [Tutorial: Hello, Quarto](https://quarto.org/docs/get-started/hello/rstudio.html)).

-   Please submit a `.zip` file that includes:

Your `app.R` file (fully working Shiny app).

A short project report (PDF).

**Reminder**: If your app fails to open or the .zip is incorrect, you will receive a score of **ZERO**. Test everything before submission.
:::

::: callout-caution
# Due Date

**Due Date:** Sunday, June 1, 2025, 11:59 PM
:::

## 2 Tasks:

### 2.1.1: Data Loading:

```{r}
library(tidyverse)
data(infert)
```

| **Variable** | **Description** | **Value** |
|-------------------------|----------------------------|-------------------|
| education | years of education | 0–5 years, 6–11 years, 12+ years |
| age | age in years | 22-24 |
| parity | number of prior live births | 1-6 |
| induced | number of prior induced abortions | 0, 1, 2 or more |
| case | case status | case, control |
| spontaneous | number of prior spontaneous abortions | 0, 1, 2 or more |
| stratum | id for each matched set or group | 1–83 |
| pooled.stratum | represent a group or broader classification of strata (may be used for conditional logistic regression) | 1 - 63 |

### 2.1.2: Data Preparation:

```{r}
infert %>% head()
```

```{r}
library(stringr)
# rename variables and remove "yrs" from education values. factorize education and case status
data(infert)
infert_ds <- infert %>%
  rename(
    Education = education,
    Age = age,
    Parity = parity,
    InducedAbortions = induced,
    SpontaneousAbortions = spontaneous,
    CaseStatus = case,
    StratumID = stratum,
    PooledStratumID = pooled.stratum
  ) %>%
  mutate(
    Education = factor(str_remove(Education, "yrs"), levels = c("0-5", "6-11", "12+")),
    CaseStatus = factor(CaseStatus, levels = c(0, 1), labels = c("control", "case"))
  )
```

```{r}
summary(infert_ds)
```

```{r}
str(infert_ds)
```

### 2.2 Shiny App Development

**User Interface (UI):**

```{r, eval = FALSE}

```

**Server Logic:**

```{r, eval = FALSE}

```

# Documentation and Presentation

## App Purpose

------------------------------------------------------------------------

## How it Works

------------------------------------------------------------------------

## Insights

------------------------------------------------------------------------

## Reflections

------------------------------------------------------------------------
