---
title: 'Mini Project 2'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author: "Instructor: Ali Abuzaid"
format: 
 pdf:
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```


:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- PHUC Lu    (pdlu)
- STUDENT 2    (NetID 2)

:::

:::{.callout-tip}
# Instructions

* This mini project is designed to give you practical experience with real-world data using R and Shiny. You’ll create an interactive web application that allows users to explore and visualize a dataset.

* Work in groups of **2 students** from the same discussion section.
* Individual submissions will not be accepted.

* * Please use the provided `MP 2.qmd` file to type your Documentation and Presentation and submit it as a PDF file. You can utilize `RStudio`for this purpose. For guidance, refer to the [Tutorial: Hello, Quarto](https://quarto.org/docs/get-started/hello/rstudio.html)).


* Please submit a `.zip` file that includes:

Your `app.R` file (fully working Shiny app).

A short project report (PDF).


**Reminder**: If your app fails to open or the .zip is incorrect, you will receive a score of **ZERO**. Test everything before submission.

:::

:::{.callout-caution}
# Due Date
**Due Date:** Sunday, June 1, 2025, 11:59 PM
:::


## 2 Tasks:

### 2.1.1: Data Loading:

```{r}
library(tidyverse)
data(infert)
```

| **Variable**  | **Description**   | **Value** |
|---------------|------------------ |-----------|
| education | years of education | 0–5 years, 6–11 years, 12+ years |
| age | age in years | 22-24 |
| parity | number of prior live births | 1-6 |
| induced | number of prior induced abortions | 0, 1, 2 or more |
|case | case status | case, control |
| spontaneous | number of prior spontaneous abortions | 0, 1, 2 or more |
| stratum | id for each matched set or group | 1–83 |
| pooled.stratum | represent a group or broader classification of strata (may be used for conditional logistic regression) | 1 - 63 |

### 2.1.2: Preparation:

```{r}
infert %>% head()
```

```{r}
library(stringr)
# remove "yrs" from education values.
infert_ds <- infert %>% mutate(education = str_remove(education, "yrs"))
infert_ds %>% head()
```

```{r}
summary(infert_ds)
```
```{r}
str(infert_ds)
```


# Documentation and Presentation

## App Purpose

----

## How it Works

----

## Insights
 
----
 
## Reflections

---
