---
title: 'Mini Project 2'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author: "Instructor: Ali Abuzaid"
format: 
 pdf:
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```

:::: callout
::: {style="text-align: center"}
[**STUDENT NAME**]{style="color: blue;"}
:::

-   Phuc Lu (pdlu)
-   Valerie De La Fuente (valeriedelafuente)
::::

::: callout-tip
# Instructions

-   This mini project is designed to give you practical experience with real-world data using R and Shiny. You’ll create an interactive web application that allows users to explore and visualize a dataset.

-   Work in groups of **2 students** from the same discussion section.

-   Individual submissions will not be accepted.

-   Please use the provided `MP 2.qmd` file to type your Documentation and Presentation and submit it as a PDF file. You can utilize `RStudio`for this purpose. For guidance, refer to the [Tutorial: Hello, Quarto](https://quarto.org/docs/get-started/hello/rstudio.html)).

-   Please submit a `.zip` file that includes:

Your `app.R` file (fully working Shiny app).

A short project report (PDF).

**Reminder**: If your app fails to open or the .zip is incorrect, you will receive a score of **ZERO**. Test everything before submission.
:::

::: callout-caution
# Due Date

**Due Date:** Sunday, June 1, 2025, 11:59 PM
:::

## 2 Tasks:

### 2.1.1: Data Loading:

```{r}
library(tidyverse)
data(infert)
```

| **Variable** | **Description** | **Value** |
|-------------------------|----------------------------|-------------------|
| education | years of education | 0–5 years, 6–11 years, 12+ years |
| age | age in years | 22-24 |
| parity | number of prior live births | 1-6 |
| induced | number of prior induced abortions | 0, 1, 2 or more |
| case | case status | case, control |
| spontaneous | number of prior spontaneous abortions | 0, 1, 2 or more |
| stratum | id for each matched set or group | 1–83 |
| pooled.stratum | represent a group or broader classification of strata (may be used for conditional logistic regression) | 1 - 63 |

### 2.1.2: Data Preparation:

```{r}
infert %>% head()
```

```{r}
library(stringr)
# rename variables and remove "yrs" from education values. factorize necessary variables
infert_ds <- infert %>%
  rename(
    Education = education,
    Age = age,
    Parity = parity,
    InducedAbortions = induced,
    SpontaneousAbortions = spontaneous,
    CaseStatus = case,
    StratumID = stratum,
    PooledStratumID = pooled.stratum
  ) %>%
  mutate(
    Education = factor(str_remove(Education, "yrs"), levels = c("0-5", "6-11", "12+")),
    CaseStatus = factor(CaseStatus, levels = c(0, 1), labels = c("control", "case")),
    Parity = factor(Parity),
    InducedAbortions = factor(InducedAbortions),
    SpontaneousAbortions = factor(SpontaneousAbortions),
    StratumID = factor(StratumID),
    PooledStratumID = factor(PooledStratumID)
  )

factor_vars <- names(infert_ds)[sapply(infert_ds, is.factor)]
numeric_vars <- names(infert_ds)[sapply(infert_ds, is.numeric)]
```

```{r}
summary(infert_ds)
```

```{r}
str(infert_ds)
```

### 2.2 Shiny App Development

**User Interface (UI):**

```{r, eval = FALSE}
# Define UI
ui <- fluidPage(
  titlePanel("Infertility Dataset Explorer"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput("variable", "Select Variable:",
                  choices = names(infert_ds)),
      
      uiOutput("plotTypeUI"),
      
      conditionalPanel(
        condition = "input.plotType == 'Histogram'",
        sliderInput("bins", "Number of Bins:", min = 5, max = 50, value = 10),
        checkboxInput("showDensity", "Overlay Density Curve", value = FALSE)
      ),
      
      conditionalPanel(
        condition = "input.plotType == 'Boxplot'",
        selectInput("boxplotGroup", "View Boxplot Across:",
                    choices = factor_vars),
        checkboxInput("showJitter", "Add Jittered Data Points", value = FALSE)
      ),
      
      checkboxInput("includeNA", "Include NA in Grouped Summary", value = FALSE),
      
      h4("Numeric Filters"),
      sliderInput("ageFilter", "Age Range:",
                  min(infert_ds$Age), max(infert_ds$Age),
                  value = c(min(infert_ds$Age), max(infert_ds$Age))),
      
      selectInput("groupBy", "Group Summary By:",
                  choices = factor_vars)
    ),
    
    mainPanel(
      plotOutput("mainPlot", hover = "plot_hover"),
      verbatimTextOutput("hoverInfo"),
      h4("Summary"),
      verbatimTextOutput("summaryOutput"),
      h4("Grouped Summary"),
      tableOutput("groupedSummary")
    )
  )
)
```

**Server Logic:**

```{r, eval = FALSE}
# Define server logic
server <- function(input, output, session) {
  
  # Dynamically adjust plot options
  output$plotTypeUI <- renderUI({
    req(input$variable)
    var <- input$variable
    
    if (is.numeric(infert_ds[[var]])) {
      selectInput("plotType", "Select Plot Type:", choices = c("Histogram", "Boxplot"))
    } else {
      selectInput("plotType", "Select Plot Type:", choices = c("Bar Chart"))
    }
  })
  
  # Reactive filtered dataset
  filtered_data <- reactive({
    infert_ds %>%
      filter(
        Age >= input$ageFilter[1], Age <= input$ageFilter[2]
      )
  })
  
  # Generate plots
  output$mainPlot <- renderPlot({
    req(input$variable, input$plotType)
    df <- filtered_data()
    var <- input$variable
    
    if (input$plotType == "Histogram") {
      p <- ggplot(df, aes_string(x = var)) +
        geom_histogram(bins = input$bins, fill = "lightpink", color = "black") +
        labs(title = paste("Histogram of", var), x = var, y = "Count") +
        theme_minimal()
      
      if (input$showDensity) {
        p <- p + geom_density(aes(y = ..count..), color = "blue", size = 1)
      }
      
      p
      
    } else if (input$plotType == "Boxplot") {
      group_var <- input$boxplotGroup
      
      p <- ggplot(df, aes_string(x = group_var, y = var)) +
        geom_boxplot(fill = "royalblue") +
        labs(title = paste("Boxplot of", var, "across", group_var), x = group_var, y = var) +
        theme_minimal()
      
      if (input$showJitter) {
        p <- p + geom_jitter(width = 0.2, alpha = 0.4)
      }
      
      p
      
    } else if (input$plotType == "Bar Chart") {
      ggplot(df, aes_string(x = var)) +
        geom_bar(fill = "lavender", color = "black") +
        labs(title = paste("Bar Chart of", var), x = var, y = "Count") +
        theme_minimal()
    }
  })
  
  # Show hover info
  output$hoverInfo <- renderPrint({
    hover <- input$plot_hover
    
    if (input$plotType == "Histogram" || input$plotType == "Boxplot") {
      if (!is.null(hover)) {
        cat("Hovered at:\n")
        cat("x:", round(hover$x, 2), "\n")
        cat("y:", round(hover$y, 2), "\n")
      } else {
        cat("Hover over the plot to see coordinates.")
      }
    } else {
      cat("Hover info disabled for bar charts (categorical x-axis).")
    }
  })
  
  # Show summary statistics
  output$summaryOutput <- renderPrint({
    req(input$variable)
    summary(filtered_data()[[input$variable]])
  })
  
  # Grouped summary
  output$groupedSummary <- renderTable({
    df <- filtered_data()
    group_var <- input$groupBy
    
    if (!input$includeNA) {
      df <- df %>% drop_na(all_of(group_var))
    }
    
    df %>%
      group_by(across(all_of(group_var))) %>%
      summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop")
  }, rownames = TRUE)
}

# Run the app
shinyApp(ui = ui, server = server)
```

# Documentation and Presentation

## App Purpose

------------------------------------------------------------------------

## How it Works

------------------------------------------------------------------------

## Insights

------------------------------------------------------------------------

## Reflections

------------------------------------------------------------------------
